---
title: "Project B for SYBB412"
author: "Emily Hsu & Shufei Fang"
date: last-modified
format: 
  html:
    toc: true
    number-sections: true
    code-fold: show
    code-tools: true
    code-overflow: wrap
    embed-resources: true
    date-format: iso
    theme: pulse
---

# R package and setup

```{r}
#| message: false
#| warning: false

knitr::opts_chunk$set(comment = NA)

library(dplyr)
library(ggplot2)
library(patchwork)
library(gt)
library(Seurat)
library(SingleR)
library(conflicted)
library(see)
library(celldex)
library(kableExtra)


library(tidyverse)

conflicts_prefer(dplyr::select(), dplyr::filter(), base::max(), stats::cor(), base::sum(), stats::quantile
                 )

theme_set(theme_lucid())

```

We decided to set a seed (123) to ensure the results are reproducible.

```{r}

set.seed(123)

```

# Load in Seurat objects from previous analysis

```{r}

raw <- readRDS("csf_merged_raw_seurat.rds")

qc  <- readRDS("csf_with_qc_metrics.rds")

```

# Quaility Control and Filtering

## QC Metric Assessement

### Violin Plot 

We start by creating a violin plot to see the distribution of our data by diagnosis group

```{r}

VlnPlot(
  qc,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  group.by = "group3",
  ncol = 3,
  pt.size = 0
) &
  geom_boxplot(width = 0.1, fill = "white", alpha = 0.5)

```

Looking at the three violin plot, we can see that most of the cells falls in a similar ranges of detected genes, total rna counts and mitochondrial percentage.

But we can also see there are some outliers, espically with mitochondrial percentage.

### Scatter Plot 

```{r}

plot1 <- FeatureScatter(qc, feature1 = "nCount_RNA", feature2 = "percent.mt")

plot2 <- FeatureScatter(qc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

```

```{r, fig.width=14, fig.height=14}

plot1 + plot2

```


### Setting up filtering parameters

Based on our previous analysis and the violin plot above, we can next choose our qc filter threshold to remove any unwanted cells.

```{r}

CSF_F <- subset(
  qc,
  subset =
    nFeature_RNA >= 300 & #removes likely low-quality droplets
    nFeature_RNA <= 3500 & #remove unsually complex cells in the upper tials (doublets or multiples)
    nCount_RNA   >= 500 & #remove cells with weak signals
    nCount_RNA   <= 15000 & #remove high outliers
    percent.mt   <= 12 #remove cells with high mt rna, normally stressed or damaged 
)

```

Than, we set the RNA assay as the default assay for downstream analsyis.

```{r}

DefaultAssay(CSF_F) <- "RNA"

```

```{r}

CSF_F <- JoinLayers(CSF_F, assay = "RNA")

```

## Normalization

```{r}

CSF_F <- NormalizeData(CSF_F)

```

Here we only select the genes that vary the most across the cells in order to help us separate cell population.

```{r}

CSF_F <- FindVariableFeatures(CSF_F, 
                              selection.method = "vst",
                              infeatures = 2000)

```

```{r}

top10 <- head(VariableFeatures(CSF_F), 10)

```

```{r}

plot1 <- VariableFeaturePlot(CSF_F)

LabelPoints(plot = plot1, points = top10, repel = TRUE)

```

## Scaling the data

Here we standardize our gene expression before perform PCA analysis.

```{r}

all.genes <- rownames(CSF_F)

CSF_F <- ScaleData(CSF_F, features = all.genes)

```

## Principal Component Analysis

```{r}

CSF_F <- RunPCA(CSF_F, features = VariableFeatures(object = CSF_F))

```

```{r}

print(CSF_F[["pca"]], dims = 1:10, nfeatures = 10)

```

PC1: myeloid vs T cells –\> edit: memory/cytotoxic T cell vs myeloid

PC2: A bit general, no specific immune lineage

PC3: cytotoxic NK/CD8 vs less-cytotoxic T cells

PC4: different myeloid subtypes

PC5: activated/helper-like T cells vs cytotoxic lymphocytes


## Determine the dimensionality of the dataset

```{r}

ElbowPlot(CSF_F, ndims = 30)

```

From the eblow plot above, I determine the dimentionnaltiy of the dataset to about 10 in order to capture the major biological structure while limiting the noise.

# Clustering and Visualization

Since we select 10 PCs as our cut off, we will first build a graph for the cells in the first 10 PCs based on their similarity.

```{r}

CSF_F <- FindNeighbors(CSF_F, dims = 1:10)

```

Next, we groups the similar cells into clusters

```{r}

CSF_F <- FindClusters(CSF_F, resolution = 0.4)

```

Finally, we make a visualization of those clusters

```{r}

CSF_F <- RunUMAP(CSF_F, dims = 1:10)

```

```{r}

p1<- DimPlot(CSF_F, reduction = "umap", label = TRUE) +
  theme(plot.title = element_blank())

```

We can see that our cell can be seperated into several distinct populations. Next, we separate everthing by Healthy vs Diseased, and Healthy vs PDD vs DLB

```{r}

p2<- DimPlot(CSF_F, reduction = "umap", group.by = "group3") +
  theme(plot.title = element_blank())

```

```{r}

p3<- DimPlot(CSF_F, reduction = "umap", group.by = "group2") +
  theme(plot.title = element_blank())

```

```{r umap-combined, fig.width=11, fig.height=13}

(p2 | p3) / (p1 | plot_spacer())+
  plot_annotation(
    title = "Visualization of Cells by Cluster and Diagnosis",
    theme = theme(
      plot.title = element_text(
        hjust = 0.5,
        size = 16,
        face = "bold"))
    )

```

# Cluster Copostition by Diagnosis Group

Next, we summarized the number and proportion of cells from each diagnosis group within each cluster

```{r}

table(Idents(CSF_F), 
      CSF_F$group3)

prop.table(table(Idents(CSF_F), 
                 CSF_F$group3), 
           margin = 1
           )
           
```

### Clearly Healthy-enriched

These clusters have a noticeably higher Healthy fraction than PDD or DLB:

Cluster 3: 88.2% Healthy

Cluster 5: 69.6% Healthy

Cluster 14: 70.1% Healthy

Cluster 0: 62.6% Healthy

Cluster 2: 59.7% Healthy

Cluster 10: 58.7% Healthy

Cluster 4: 50.6% Healthy

So yes, cluster 3 is especially strong, and 5 and 14 are clearly Healthy-enriched

### Clearly PDD-enriched

These clusters lean more toward PDD:

Cluster 1: 51.2% PDD Cluster 4: 49.4% PDD Cluster 6: 88.9% PDD Cluster 7: 79.3% PDD Cluster 8: 60.6% PDD Cluster 11: 64.9% PDD Cluster 12: 53.6% PDD Cluster 13 64.9%

So one correction to your earlier impression: cluster 13 is actually more PDD-enriched than Healthy-enriched.

### DLB

DLB is low in most clusters, but that is hard to interpret because you only have 2 DLB samples. The only cluster where DLB is really notable is:

Cluster 1: 26.2% DLB

That is worth noticing, but I would be very cautious about making a strong DLB claim yet

## Quantify cluster composition by sample

```{r}

table(Idents(CSF_F), 
      CSF_F$sample_title)

prop.table(table(Idents(CSF_F),
                 CSF_F$sample_title), 
           margin = 1
           )

```

# Annotation

First we set up to find the marker gene for every clusters

```{r}

markers <- FindAllMarkers(
  CSF_F,
  only.pos = TRUE, #Only kept the upregulated gene
  min.pct = 0.25, #Must be detected in at least 25% of cells
  logfc.threshold = 0.25
)

```

Next, we select the 10 strongest marker gene for each cluster

```{r}

top_m <- markers|>
  group_by(cluster)|>
  slice_max(order_by = avg_log2FC, n =10) #sort by average log2 fold change

```

```{r}

print(top_m, n = 200)

```

### Using SingleR for cell type annotation

```{r}

sce <- as.SingleCellExperiment(CSF_F)

```

```{r}

library(celldex)

ref <- MonacoImmuneData()

```

```{r}

main <- SingleR(
  test = sce,
  ref = ref,
  labels = ref$label.main
)

```

```{r}

CSF_F$SingleR_main <- main$labels

CSF_F$SingleR_main_pruned <- main$pruned.labels

```

```{r, fig.width=11, fig.height=13}

DimPlot(CSF_F, group.by = "SingleR_main_pruned", label = TRUE, repel = TRUE)

```

```{r, fig.width=11, fig.height=13}

table(CSF_F$seurat_clusters, CSF_F$SingleR_main_pruned)

prop.table(
  table(CSF_F$seurat_clusters, CSF_F$SingleR_main_pruned),
  margin = 1
)

```

Cluster 0: activated CD4+ T cells, likely effector/activated helper T cells

Cluster 1: activated CD4+ T cells, effector/memory-activated state

Cluster 2: cytotoxic CD8+ T cells, effector/activated CTL state

Cluster 3: memory CD4+ T cells

Cluster 4: activated effector memory CD8+ T cells/NK-like cytotoxic T cells

Cluster 5: myeloid/monocyte-derived dendritic cells

Cluster 6: KLRB1+ memory/effector CD4+ T cells

Cluster 7: macrophages

Cluster 8: NK cells

Cluster 9: inflammatory monocytes

Cluster 10: NK cells

Cluster 11: dendritic cells (cDC2 dendritic cells)

Cluster 12: Mixed

Cluster 13: non-immune

Cluster 14: plasmacytoid dendritic cell (pDC)

Just to make us see it better, we will relabel all the clusters with biological names

```{r}

Idents(CSF_F) <- "seurat_clusters"

p <- DimPlot(
  CSF_F,
  reduction = "umap",
  label = FALSE
) + theme(legend.position = "right")

cluster_map <- c(
    "0" = "Act_T",        # 0
    "1" = "Mem_T",        # 1
    "2" = "Act_T2",       # 2
    "3" = "CD8_T",        # 3
    "4" = "Cyt_NK",       # 4
    "5" = "KLRB1_T",      # 5
    "6" = "CD1C_DC",      # 6
    "7" = "Macro",        # 7
    "8" = "NK",       # 8
    "9" = "Mono_inflam",  # 9
    "10" = "NK2",     # 10
    "11" = "cDC2_DC2",    # 11
    "12" = "Mixed",        # 12
    "13" = "Non-immune",    # 13
    "14" = "pDC_like"      # 14
)

centers <- Embeddings(CSF_F, "umap") |>
  as.data.frame() |>
  rownames_to_column("cell") |>
  left_join(
    CSF_F[[]] |>
      rownames_to_column("cell") |>
      select(cell, seurat_clusters),
    by = "cell"
  ) |>
  group_by(seurat_clusters) |>
  summarise(
    umap_1 = median(umap_1),
    umap_2 = median(umap_2),
    .groups = "drop"
  ) |>
  mutate(label = cluster_map[as.character(seurat_clusters)])

```

```{r}

p4 <- p +
  geom_text(
    data = centers,
    aes(x = umap_1, y = umap_2, label = label),
    inherit.aes = FALSE,
    size = 3.5
  )

```

Combine the plot for a presentable visualization

```{r umap-combined, fig.width=11, fig.height=13}

combined_umap <- (p2 | p3) / (p4 | plot_spacer())+
  plot_annotation(
    title = "Visualization of Cells by Cluster and Diagnosis",
    theme = theme(
      plot.title = element_text(
        hjust = 0.5,
        size = 16,
        face = "bold"))
    )

```

```{r}
ggsave(
  filename = "combined_umap_highres.tiff",
  plot = combined_umap,
  width = 10,
  height = 12,
  units = "in",
  dpi = 600,
  compression = "lzw"
)
```

# Save the annotated whole CSF object

```{r}

saveRDS(csf_filtered, "annotated_4_15_26.rds")

```

# Select subset

Cluster 3 is a strong CD8/cytotoxic T-cell cluster: CD8A, CD8B, GZMH, CCL4, and NKG7 are all classic cytotoxic lymphocyte markers

Cluster 4 Clusters 8 and 10 are all cytotoxic lymphoctye program

```{r}
# 2 options:
    # Cluster 3,
```

# Including 3 and 4

```{r}

Idents(CSF_F) <- CSF_F$seurat_clusters

```

```{r}

cd8_c34 <- subset(CSF_F, idents = c("3","4"))

```

## Return to Seurat workflow for Cluster 3&4

```{r}

table(Idents(cd8_c34))

```

Normalization

```{r}

cd8_c34 <- NormalizeData(cd8_c34)

```

```{r}

DefaultAssay(cd8_c34) <- "RNA"

cd8_c34[["RNA"]]$scale.data <- NULL

```

Indentify the most informative gene

```{r}

cd8_c34 <- FindVariableFeatures(cd8_c34, 
                               selection.method = "vst", 
                               nfeatures = 2000)

```

Center and standerdize

```{r}

cd8_c34 <- ScaleData(cd8_c34)

```

PCA

```{r}

cd8_c34 <- RunPCA(cd8_c34, npcs = 30)

```

```{r}

ElbowPlot(cd8_c34, ndims = 30)

```

```{r}

cd8_c34 <- FindNeighbors(cd8_c34, dims = 1:10)

```

```{r}

cd8_c34 <- FindClusters(cd8_c34, resolution = 0.3)

```

```{r}

cd8_c34 <- RunUMAP(cd8_c34, dims = 1:8)

```

Visualization

```{r}

DimPlot(cd8_c34, reduction = "umap", label = TRUE)

```

Healthy vs LBD

```{r}

DimPlot(cd8_c34, reduction = "umap", group.by = "group2")

```

`Can see a mostly clear right vs left.` Showing there is a diagnositic difference

Healthy vs DLB vs PDD

```{r}

DimPlot(cd8_c34, reduction = "umap", group.by = "group3")

```

Right side looked mostly health, left side looked mostly PDD. Bottom left has a DLB cluster

```{r}
table(Idents(cd8_c34), cd8_c34$SingleR_main_pruned)

prop.table(
  table(Idents(cd8_c34), cd8_c34$SingleR_main_pruned),
  margin = 1
)
```

Since our analysis are CD8+ T cells focused, I will select subcluster 0, 1, 2 for furthur anlysis.

Quantify each group

```{r}

table(Idents(cd8_c34), cd8_c34$group2)

prop.table(table(Idents(cd8_c34), cd8_c34$group2), margin = 1)

```

Subcluster 1 is strongly health associated.

Subcluster 2 is strongly LBD associated

```{r}

table(Idents(cd8_c34), cd8_c34$group3)

prop.table(table(Idents(cd8_c34), cd8_c34$group3), margin = 1)

```

Subcluster 1 is strongly health associated.

Subcluster 2 is strongly diseased with skew toward PDD.

There isn't a DLB dominant group.

# Focuse on subcluster 1 and 2

```{r}

cd8_c34 <- JoinLayers(cd8_c34, assay = "RNA")

```

## Maker indentification

```{r}

cd8_c34m <- FindAllMarkers(
  cd8_c34,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

```

```{r}

cd8_c34m_t <- cd8_c34m |>
  group_by(cluster) |>
  slice_max(order_by = avg_log2FC, n =10)

```

```{r}

cd8_c34m_t |>
  dplyr::select(cluster, gene, avg_log2FC, pct.1, pct.2, p_val_adj) |>
  dplyr::arrange(cluster, dplyr::desc(avg_log2FC)) |>
  print(n = Inf, width = Inf)

```

### Using SingleR

```{r}
sce_cd8 <- as.SingleCellExperiment(cd8_c34)

if (!"logcounts" %in% assayNames(sce_cd8)) {
  sce_cd8 <- logNormCounts(sce_cd8)
}

ref <- MonacoImmuneData()

pred_main_cd8 <- SingleR(
  test = sce_cd8,
  ref = ref,
  labels = ref$label.main
)

pred_fine_cd8 <- SingleR(
  test = sce_cd8,
  ref = ref,
  labels = ref$label.fine
)

cd8_c34$SingleR_main <- pred_main_cd8$labels
cd8_c34$SingleR_main_pruned <- pred_main_cd8$pruned.labels

cd8_c34$SingleR_fine <- pred_fine_cd8$labels
cd8_c34$SingleR_fine_pruned <- pred_fine_cd8$pruned.labels
```

```{r}
table(Idents(cd8_c34), cd8_c34$SingleR_main_pruned)

prop.table(
  table(Idents(cd8_c34), cd8_c34$SingleR_main_pruned),
  margin = 1
)
```

The strongest contrast is subcluster 1 and 2

#Continue to Subcluster for 0 & 1 & 2

```{r}

Idents(cd8_c34) <- "seurat_clusters"

```

```{r}

cd8_012<- subset(cd8_c34, idents = c("0", "1", "2"))

```

```{r}

DefaultAssay(cd8_012) <- "RNA"

cd8_012[["RNA"]]$scale.data <- NULL

```

```{r}

cd8_012 <- NormalizeData(cd8_012)

```

```{r}

cd8_012 <- FindVariableFeatures(
  cd8_012,
  selection.method = "vst",
  nfeatures = 2000
)

```

```{r}

cd8_012 <- ScaleData(cd8_012)

```

```{r}

cd8_012 <- RunPCA(cd8_012, npcs = 30)

```

```{r}

ElbowPlot(cd8_012, ndims = 30)

```

```{r}
cd8_012 <- FindNeighbors(cd8_012, dims = 1:10)
cd8_012 <- FindClusters(cd8_012, resolution = 0.2)
cd8_012 <- RunUMAP(cd8_012, dims = 1:10)

DimPlot(cd8_012, reduction = "umap", label = TRUE)
DimPlot(cd8_012, reduction = "umap", group.by = "group2")
DimPlot(cd8_012, reduction = "umap", group.by = "group3")
```

```{r}
table(Idents(cd8_012), cd8_012$group2)
prop.table(table(Idents(cd8_012), cd8_012$group2), margin = 1)

table(Idents(cd8_012), cd8_012$group3)
prop.table(table(Idents(cd8_012), cd8_012$group3), margin = 1)
```

```{r}

markers_cd8_012 <- FindAllMarkers(
  cd8_012,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

```

```{r}
markers_cd8_012 |>
  dplyr::select(cluster, gene, avg_log2FC, pct.1, pct.2, p_val_adj) |>
  dplyr::arrange(cluster, dplyr::desc(avg_log2FC)) |>
  dplyr::group_by(cluster) |>
  dplyr::slice_head(n = 10) |>
  print(n = Inf, width = Inf)
```

```{r}

Idents(cd8_012) <- "seurat_clusters"

cd8_final <- subset(cd8_012, idents = c("1", "2"))

```

```{r}
umap_df <- as.data.frame(Embeddings(cd8_final, "umap"))
colnames(umap_df) <- c("UMAP_1", "UMAP_2")
umap_df$cluster <- as.character(Idents(cd8_final))

centers <- umap_df %>%
  group_by(cluster) %>%
  summarize(x = median(UMAP_1), y = median(UMAP_2)) %>%
  mutate(label = case_when(
    cluster == "1" ~ "Healthy CD8+ T cells",
    cluster == "2" ~ "Diseased CD8+ T cells",
    TRUE ~ cluster
  ))

```


```{r}
p6 <- DimPlot(cd8_final, reduction = "umap", label = FALSE) +
  theme(plot.title = element_blank()) +
  geom_text(data = centers,
            aes(x = x, y = y, label = label),
            size = 3, color = "black",
            inherit.aes = FALSE) +
  expand_limits(x = c(min(umap_df$UMAP_1) - 2, max(umap_df$UMAP_1) + 2))

p7<- DimPlot(cd8_final, reduction = "umap", group.by = "group2")+
  theme(plot.title = element_blank())

p8<- DimPlot(cd8_final, reduction = "umap", group.by = "group3")+
  theme(plot.title = element_blank())

table(Idents(cd8_final), cd8_final$group2)
prop.table(table(Idents(cd8_final), cd8_final$group2), margin = 1)

table(Idents(cd8_final), cd8_final$group3)
prop.table(table(Idents(cd8_final), cd8_final$group3), margin = 1)
```

```{r umap-combined, fig.width=11, fig.height=13}

final_plot <-(p8 | p7) / (p6 | plot_spacer()) +
  plot_annotation(
    title = "Visualization of CD8+ T Cells by Cluster and Diagnosis",
    theme = theme(
      plot.title = element_text(
        hjust = 0.5,
        size = 16,
        face = "bold"
      )
    )
  )

```

```{r, fig.width=11, fig.height=13}
final_plot
```

```{r}
ggsave("cd8_final.png", plot = final_plot, width = 12, height = 8, dpi = 300)

ggsave("cd8_final.tiff", plot = final_plot,
       width = 12, height = 8, dpi = 300, compression = "lzw")
```

##Fiinal Maker Anlysis

```{r}

markers_cd8_final <- FindAllMarkers(
  cd8_final,
  only.pos = TRUE,
  min.pct = 0.25,
  logfc.threshold = 0.25
)

```

```{r}

top_markers <- FindAllMarkers(cd8_final, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25) |>
  mutate(cluster = as.character(cluster)) |>
  group_by(cluster) |>
  slice_max(avg_log2FC, n = 5) |>
  summarise(Markers = paste(gene, collapse = "<br>"), .groups = "drop")

```

```{r}

diag_counts <- tibble(
  cluster = as.character(Idents(cd8_final)),
  group2  = as.character(cd8_final$group2)
) |>
  dplyr::count(cluster, group2) |>
  group_by(cluster) |>
  mutate(value = sprintf("%d (%.1f%%)", n, 100 * n / sum(n))) |>
  ungroup() |>
  pivot_wider(id_cols = cluster, names_from = group2, values_from = value, values_fill = "0 (0.0%)")

```

#### Final table

```{r}
final_tbl <- final_tbl %>%
  mutate(State = case_when(
    Cluster == "1" ~ "Healthy CD8+ T cells",
    Cluster == "2" ~ "Diseased CD8+ T cells"
  )) %>%
  select(Cluster, State, Markers, Healthy, LBD) %>%
  mutate(Markers = sapply(Markers, function(x) paste(x, collapse = ", ")))

```

```{r}
final_tbl %>%
  kbl(escape = FALSE, 
      caption = "<div style='text-align:center; font-size:20px; font-weight:bold; color:black;'>CD8 Subcluster Marker Summary</div>",
      align = c("c", "l", "l", "c", "c")) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, background = "#e8e8e8")
```


# Session Info

```{r}

sessionInfo()

```
